Trap spaces of Boolean networks are conflict-free siphons of their Petri net encoding

### 2023-04-11 manuscripts/TCS2023/tcs2023.tex
Some small changes

@@

\begin{example}
We give a Boolean network \(\mathcal{N} = (V, F)\), where \(V = (x_1, x_2)\) and \(F = (f_1, f_2)\) with \(f_1 = (x_1 \land x_2) \lor (\neg x_1 \land \neg x_2), f_2 = (x_1 \land x_2) \lor (\neg x_1 \land \neg x_2)\). Herein, \(\land\), \(\lor\), and \(\neg\) denote the conjunction, disjunction, and negation logical operators, respectively.\label{example:BN}
\end{example}

@@

Note that transforming a Boolean network defined by its functions into its Petri net encoding roughly relies on obtaining conditions for the activation and inactivation of the states. In~\cite{chaouiya2004qualitative} this took the form of the whole truth table of the Boolean functions, but as shown in Appendix 1 of~\cite{chatain2014characterization} computing Disjunctive Normal Forms (DNF) of each Boolean function is enough.
Though this might appear quite computationally intensive it is important to remark first that contrary to the prime-implicants case, there is no need to find \emph{minimal} DNFs.
One way to look at this is to consider that this amounts to a similar approach as that used in~\cite{DBLP:conf/ictai/ChevalierFPZ19} but with the encoding of both activation and inhibition functions as DNFs in order to take into account possible non-local-monotonicity.
This does not change the worst-case-complexity (obtaining a single DNF being exponential) but might matter a lot in practice.
As such, we will explore how this transformation, here using BDDs in \texttt{bioLQM} or directly in our tool using the \texttt{pyeda}\footnote{\url{https://pyeda.readthedocs.io/en/latest/}} library, and the one based on the most-permissive semantics compare in Section~\ref{sec:eval} on evaluation.

@@

Intuitively, a siphon is a set of places that once unmarked remains so.
If it is conflict-free then its dual corresponds to a partial-state of the model such that whatever update, the fixed values remain so (since the unmarked places remain unmarked).
This is precisely the definition of a trap space and maximality of the siphon is equivalent to as many fixed values as possible, hence minimality of the trap space.
For example, the Boolean network given in Example~\ref{example:BN} has two trap spaces, \(m_1 = 11\) and \(m_2 = \star\star\).
The Petri net encoding of this Boolean network has five generic siphons, \(S_1 = \emptyset\), \(S_2 = \{p_{x_1}, \overline{p}_{x_1}\}\), \(S_3 = \{p_{x_2}, \overline{p}_{x_2}\}\), \(S_4 = \{\overline{p}_{x_1}, \overline{p}_{x_2}\}\), and \(S_5 = \{p_{x_1}, \overline{p}_{x_1}, p_{x_2}, \overline{p}_{x_2}\}\).
However, only \(S_1\) and \(S_4\) are conflict-free siphons and correspond to \(m_2\) and \(m_1\), respectively.
Since \(S_1 \subset S_4\), \(S_4\) is a maximal siphon corresponding to the minimal trap space \(m_1\).
Hereafter, we formally prove that a (maximal) conflict-free siphon is equivalent to a (minimal) trap space.

>>>>

\newcommand{\revBelaid}[1]{\textcolor{red}{#1}}

@@

\begin{example}
We give a Boolean network \(\mathcal{N} = (V, F)\), where \(V = (x_1, x_2)\) and \(F = (f_1, f_2)\) with \(f_1 = (x_1 \land x_2) \lor (\neg x_1 \land \neg x_2), f_2 = (x_1 \land x_2) \lor (\neg x_1 \land \neg x_2)\). Herein, \(\land\), \(\lor\), and \(\neg\) denote the \revBelaid{logical conjunction, disjunction, and negation operators}, respectively.\label{example:BN}
\end{example}

@@

Note that transforming a Boolean network defined by its functions into its Petri net encoding roughly relies on obtaining conditions for the activation and inactivation of the states. In~\cite{chaouiya2004qualitative} this took the form of the whole truth table of the Boolean functions, but as shown in Appendix 1 of~\cite{chatain2014characterization} computing Disjunctive Normal Forms (DNF) of each Boolean function is enough.
Though this might appear quite computationally intensive it is important to remark first that contrary to the prime-implicants case, there is no need to find \emph{minimal} DNFs.
One way to look at this is to consider that this amounts to a similar approach as that used in~\cite{DBLP:conf/ictai/ChevalierFPZ19} but with the encoding of both activation and inhibition functions as DNFs in order to take into account possible non-local-monotonicity.
This does not change the worst-case-complexity (obtaining a single DNF being exponential) but might matter a lot in practice.
As such, we will explore how this transformation, here using BDDs in \texttt{bioLQM} or directly in our tool using the \texttt{pyeda}\footnote{\url{https://pyeda.readthedocs.io/en/latest/}} library, and the one based on the most-permissive semantics \revBelaid{compare with each other in Section~\ref{sec:eval}}.

@@

Intuitively, a siphon is a set of places that once unmarked remains so.
If it is conflict-free then its dual corresponds to a \revBelaid{subspace} of the model such that whatever update, the fixed values remain so (since the unmarked places remain unmarked).
This is precisely the definition of a trap space and maximality of the siphon is equivalent to as many fixed values as possible, hence minimality of the trap space.
For example, the Boolean network given in Example~\ref{example:BN} has two trap spaces, \(m_1 = 11\) and \(m_2 = \star\star\).
The Petri net encoding of this Boolean network has five generic siphons, \(S_1 = \emptyset\), \(S_2 = \{p_{x_1}, \overline{p}_{x_1}\}\), \(S_3 = \{p_{x_2}, \overline{p}_{x_2}\}\), \(S_4 = \{\overline{p}_{x_1}, \overline{p}_{x_2}\}\), and \(S_5 = \{p_{x_1}, \overline{p}_{x_1}, p_{x_2}, \overline{p}_{x_2}\}\).
However, only \(S_1\) and \(S_4\) are conflict-free siphons and correspond to \(m_2\) and \(m_1\), respectively.
Since \(S_1 \subset S_4\), \(S_4\) is a maximal siphon corresponding to the minimal trap space \(m_1\).
Hereafter, we formally prove that a (maximal) conflict-free siphon is equivalent to a (minimal) trap space.
